Complete Liver Model for Real-time Simulation

### 2013-02-27 Doc/Miccai2013/miccai2013.tex
more on the miccai paper
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@517 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\usepackage{makeidx}  % allows for indexgeneration

@@

\begin{document}
%
%\frontmatter          % for the preliminaries
%
\mainmatter              % start of the contributions
%
\title{TODO: Title} % TODO
%
\titlerunning{TODO: Title}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Anonymous}
%\author{%
%Tom\'a\v{s} Golembiovsk\'y\inst{1,2} \and%
%Igor Peterl\'ik\inst{3} \and%
%%Christian Duriez\inst{2} \and%
%Stephan\'e Cotin\inst{2,3}%
%}
%
%\authorrunning{Tom\'a\vs Golembiovsk\'y et al.} % abbreviated author list (for running head)
%
%\institute{%
%Faculty of Informatics, Masaryk University, Brno, Czech Republic\and%
%INRIA Lille -- Nord, France\and%
%IHU, Strasbourg, France%
%}
%
\maketitle

% TODO
\begin{abstract}
This work provides a complete physical model for the deformations of the liver with its three constituents: parenchyma, vessels and the Glisson's capsule.
Some recent experimental results \cite{Ahn2010} have shown that there is a significant difference of stiffness between the Glisson's capsule and the parenchyma. 
Moreover, it has been observed that the capsule plays an important role for the deformation at the local level \cite{Hollenstein2006}.
However, the Glisson's capsule is very thin and its influence for the global deformations of the liver was never highlighted.
In this work, we propose to use a membrane FEM model for the capsule, that is coupled with a vascularized model of the liver.
We show that the approach is able to reproduce the deformations observed at the local level.
Additionally, we show, by simulating natural global deformations of the liver observed on real images (due to gravity), that the results have important differences with and without the capsule.

@@

We propose to further extend the tetrahedral based model of the parenchyma
with vascularized model \cite{Peterlik2012} by the layer of membrane elements.
(Modelling the capsule with tetrahedral elements is unfeasible -- many
elements or largely deformed elements.)

Based on the small thickness it seems reasonable to assume that no bending
strains limmit the deformation of the capsule. Therefore modelling the
capsule with membrane element only should be adequate.

@@

@@

\subsection{parenchyma} %{{{

\TG{TODO: write something like this} % (src: Peterlik2012)
\begin{quote}
While most studies concerning the material properties of liver parenchyma
agree on a viscoelastic behaviour (see [5] or [7] for instance), we employ
a simpler non-linear elastic model. This is essentially motivated by the
fact that we are not focusing on the transient part of the deformation but
rather the static equilibrium under some specific loading conditions.
\end{quote}

We model the parenchyma using corotational finite elmets. The corotational
formulation allows us to deal with geometrical non-linearities but still
rely on linear stress-strain relationship. Thus rotations and large
deformations are possible. In the corotational formulation the deformation
of each element is expressed in the local frame of reference. There are
several ways of computing the corotational frame for elements, we rely on
the geometrical method proposed in \cite{Nesme2005}.
% NOTE: This is very simplified and could be extended.
The stiffness matrix $\Mat{K}$ can be expressed in terms of the deformation
$u$ and the force as $\Vec{f} = \Mat{K}(\Vec{u}) \Vec{u}$. The resulting
system of linear equations is then solved by direct LDL solver.

@@

While in general we could use arbitrary surface mesh for a capsule we use
the fact that the parenchyma is modelled by tetrahedral elements which have
triangular faces. Because the boundary of the volumetric mesh is already
triangulated, we only extract the boundary triangles which we use for the
model of capsule.

@@

\section{Experiments/Evaluation}

setup:

@@

Test 2) Indentation test
  * perform indentation and analyze the profile
  * show how stiffer membrane affect the result

\section{Results}

@@

>>>>

\usepackage{makeidx}
\usepackage[pdftex]{graphicx}

@@

%\Figure{width}{name}{caption}
\newcommand{\Figure}[3]{%
\begin{figure}[htb]
  \centering
  \includegraphics[width=#1]{#2}
  \caption{\label{fig-#2}#3}
\end{figure}}

\begin{document}
%
%\frontmatter          % for the preliminaries
%
\mainmatter              % start of the contributions
%
\title{Complete Liver Model for Real-time Simulation} % TODO
% Or: Modelling the Glisson's Gapsule in Real-time Simulation of Liver
% or anything else ...
%
\titlerunning{TODO: Title}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Anonymous}
%\author{%
%Tom\'a\v{s} Golembiovsk\'y\inst{1,2} \and%
%Igor Peterl\'ik\inst{3} \and%
%%Christian Duriez\inst{2} \and%
%Stephan\'e Cotin\inst{2,3}%
%}
%
%\authorrunning{Tom\'a\vs Golembiovsk\'y et al.} % abbreviated author list (for running head)
%
%\institute{%
%Faculty of Informatics, Masaryk University, Brno, Czech Republic\and%
%INRIA Lille -- Nord, France\and%
%IHU, Strasbourg, France%
%}
%
\maketitle

\begin{abstract}
This work provides a complete physical model for the deformations of the liver with its three constituents: parenchyma, vessels and the Glisson's capsule.
Some recent experimental results \cite{Ahn2010} have shown that there is a significant difference of stiffness between the Glisson's capsule and the parenchyma. 
Moreover, it has been observed that the capsule plays an important role for the deformation at the local level \cite{Hollenstein2006}.
However, the Glisson's capsule is very thin and its influence for the global deformations of the liver was never highlighted.
In this work, we propose to use a membrane FEM model for the capsule, that is coupled with a vascularized model of the liver.
We show that the approach is able to reproduce the deformations observed at the local level.
Additionally, we show, by simulating natural global deformations of the liver observed on real images (due to gravity), that the results have important differences with and without the capsule.

@@

We propose to further extend the tetrahedral based model of the parenchyma
with vascularized model \cite{Peterlik2012} by the layer of membrane elements.
(Modeling the capsule with tetrahedral elements is unfeasible -- many
elements or largely deformed elements.)

Based on the small thickness it seems reasonable to assume that no bending
strains limmit the deformation of the capsule. Therefore modeling the
capsule with membrane element only should be adequate.

@@

% -----

The liver is usually modeled as a homogeneous isotropic (visco-)elastic
material \TG{need refs.?}. It is however non-homogeneous and modeling it as
such necessarily requires complex non-linear material formulation. Peterlik
et al. \cite{Peterlik2012} showed the importance of the vascular structures
on the liver deformation.

Another overlooked component is the Glisson's capsule. Despite it's very
small thickness it has high elastic modulus that is in three orders of
magnitude larger than that of the parenchyma. As such it plays an important
role in both global and local deformations of the liver.

It is not possible to model such thin structure with classical tetrahedral
elements.

% -----

@@

\subsection{Parenchyma} %{{{

It is known that the parenchyma exhibits viscoealstic behaviour \TG{Cite
something}, we however employ simpler elastic model as we are not so much
interested in time-dependent behaviour but rather in static equilibrium
under certain conditions.
%
The parenchyma is modeled using corotational finite elmets. The corotational
formulation allows us to deal with geometrical non-linearities but still
rely on linear stress-strain relationship. Thus rotations and large
deformations are possible. In the corotational formulation the deformation
of each element is expressed in the local frame of reference. There are
several ways of computing the corotational frame for elements, we rely on
the geometrical method proposed in \cite{Nesme2005}.
% NOTE: This description of corotational method is very simplified and could be extended.
The stiffness matrix $\Mat{K}$ can be expressed in terms of the deformation
$u$ and the force as $\Vec{f} = \Mat{K}(\Vec{u}) \Vec{u}$. The resulting
system of linear equations is then solved by direct LDL solver.

@@

While in general we could use arbitrary surface mesh for a capsule we use
the fact that the parenchyma is modeled by tetrahedral elements which have
triangular faces. Because the boundary of the volumetric mesh is already
triangulated, we only extract the boundary triangles which we use for the
model of capsule.

@@

\section{Experiments} %{{{

To evaluate the model we have implemented in the
SOFA\footnote{www.sofa-framework.org} framework and performed a set of
numerical simulations. We have tests the influence of the capsule for both
local and global deformations. In the simulations material properties for a
porcine liver were used as reported by Umale \cite{Umale2013}. The
parenchyma was modeled with elastic modulus 1980~Pa and the capsule has
elastic modulus 8~MPa and thickness 20~$\mu$m.
\TG{The vessels have same parameters as Igor used in his work. Is this OK
or should we use values reported by Sagar for consistency?}

\subsection{Local deformations}

Hollenstain et al. \cite{Hollenstein2006} measured the response of the
bovine liver to the negative pressure using their aspiration test. One of
the observations was that the Glissons capsule has non-negligible influence
and modeling only the parenchyma leads to overestimation of the
deformation. We reproduced the test with our model and acquired similar
results.

The aspiration device consists of a tube with 1 cm in diameter and is able
to control internal pressure in the tube. The test is performed by
attaching the tube to the tissue and meassuring the tissue response. We
have set up a simulation in the spirit of this test. 
% TODO: fix the scene! tube has 5 mm radius, not 1 cm
The simulation is performed with a 3x3x3 cm cube of liver tissue and tube
with 1 cm radius. The pressure of 2 kPa is simulated inside the tube.

Figure \ref{fig-aspiration} shows the profiles of cuts in the middle of the
cube for the model with and without capsule. It can be seen that the
deformation for the model without capsule is much larger which is in
aggreement with the data obtained by Hollenstain et al.

\Figure{4in}{aspiration}{TODO: aspiration test; fix figure, add screenshot}

\subsection{Global Deformations}

@@

@@

- we have proposed a complete liver model that employs membrane element to model
the capsule and extends work by Peterlik et al.

- we have reproduced aspiration test, results are in aggreement with Hollenstain
  et al.

- we have shown influence on global deformations
